Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-18 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

Outline of revisions:
1. generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal
2. expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation
3. changed discussion and plots to involve branch lengths and not fidelities

@@

Some of the reviewers have expressed their opinion that we reduce our emphasis on mathematical development in exchange for a more numerical-optimization-based approach.
Our primary goal was to mathematically prove inconsistency of joint inference, which is what we have done.
As pointed out by reviewer 4, this answers an open question, which has special relevance given the recent work of Sagulenko et al.
We provide some numerical results to confirm our derivations and to provide a comparison.

As we now emphasize, this work is not the final word on the joint inference question from an applied perspective.
%das: added
%Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
%Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
An investigation of these aspects deserves another paper.

@@

> Finally, please pay VERY careful attention to improving your
> literature review (reviewer 4) and making a very clear case about how
> your work IMPROVES and EXTENDS previous results. And, please, do get
> rid of the Farris zone. The name "Farris" should not be attached to
> results that have to do with ML estimation, given the long and
> torturous efforts of cladists to undermine the entire line of inquiry.

We are grateful to reviewer 4 for pointing us to literature of which we were not aware.
We have greatly expanded on the discussion in Goldman (1990) about the limitations of joint optimization of phylogenetic parameters and ancestral states.
The previous work concerns an example where branch lengths are fixed and equal, while here we show cases where inconsistency arises when branch lengths are treated as unknown parameters.
Additionally, we have placed joint optimization and maximum likelihood on level footing by viewing them both as optimization procedures over topologies and branch lengths.
Goldman (1990) does not compare maximum likelihood and maximum relative likelihood as the latter assumes equal branch lengths while the former does not.
We have added this to our discussion.

@@

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).
>
> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.
>
> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3 in the updated paper).
We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree.
Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
For reasonable parameter settings, one could investigate the error accumulation given fixed ancestral states, but this was not our goal and will require another paper.

### 2018-07-18 joint_inf.tex
go through TODOs and responses

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
%TODO: double check the below sentence
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3.
\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.

@@

While we have shown inconsistency in topology estimation and investigated the performance of branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

>>>>

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3, though we show not all branches need to be long for inconsistency to arise.
\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.

@@

While we have shown inconsistency in topology estimation and investigated the performance of branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide empirical estimates for regions of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) beg further investigation.
Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
Further study is needed to determine the impact of these results on problems that are of interest to practitioners.
Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.
